# Data Science Capstone Project

## Synopsis
The data for the project is a corpus of three documents containing not formatted text written in English. 

The corpora have been crawled from the three different public web sources: Twitter, blogs and news articles.


## Data acquisition

**Load libraries and functions**
```{r, echo = FALSE, chache = TRUE, warning = FALSE, message = FALSE} 
source("load.R")
load.libraries.functions()
```

## Exploratory analysis

### Basic summaries
**Twitter file**
```{r, echo = FALSE, chache = TRUE} 
basic.summaries.file("twitter.txt")
```
**Blog file**
```{r, echo = FALSE, chache = TRUE} 
basic.summaries.file("blogs.txt")
```
**News file**
```{r, echo = FALSE, chache = TRUE} 
basic.summaries.file("news.txt")
```

**Ngram distribution**

```{r, echo=FALSE, chache=TRUE, warning=FALSE, message=FALSE, fig.height=5, fig.width=10} 
size <- 50
tdm.list.twitter <- prepare.corpus("twitter.txt", size)
par(mar=c(5.1,10,4.1,2.1), mfrow=c(1,3), oma = c(0, 0, 2, 0))
for (n in 1:3) display.barplot(tdm.list.twitter[[n]], n, "yellow")
mtext("Twitter", outer = TRUE, cex = 1.5)
tdm.list.blogs <- prepare.corpus("blogs.txt", size)
for (n in 1:3) display.barplot(tdm.list.blogs[[n]], n, "green")
mtext("Blog", outer = TRUE, cex = 1.5)
tdm.list.news <- prepare.corpus("news.txt", size)
for (n in 1:3) display.barplot(tdm.list.news[[n]], n, "purple")
mtext("News", outer = TRUE, cex = 1.5)
```


**Unigram Clouds**

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=20}
l <- list(tdm.list.twitter[[1]], tdm.list.blogs[[1]], tdm.list.news[[1]])
display.clouds(l)
```



**Style analysis**

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}  
res <- stylo()
```

# Appendix

### Summary statistics 
**Twitter file**
```{r, echo = FALSE, chache = TRUE, warning = FALSE, message = FALSE}  
for (n in 1:4) display.summaries(tdm.list.twitter[[n]], n)
```   
**Blog file**
```{r, echo = FALSE, chache = TRUE, warning = FALSE, message = FALSE}  
for (n in 1:4) display.summaries(tdm.list.blogs[[n]], n)
``` 
**News file**
```{r, echo = FALSE, chache = TRUE, warning = FALSE, message = FALSE}  
for (n in 1:4) display.summaries(tdm.list.news [[n]], n)
``` 
