### Executive summary

**Problem description**  
You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

- Is an automatic or manual transmission better for MPG?
- Quantify the MPG difference between automatic and manual transmissions.

**The data**  
The data for this analysis is the data set mtcars of the R dataset package. It was extracted from the 1974 Motor Trend US magazine and it contains characteristics like fuel consumption, automobile design and performance for 32 automobiles. The complete list of variables is:   

Name  | Description
------| -------------
mpg   | Miles/(US) gallon
cyl   | Number of cylinders
disp  | Displacement (cu.in.)
hp    | Gross horsepower
drat  | Rear axle ratio
wt    | Weight (lb/1000)
qsec  | 1/4 mile time
vs    | V/S
am    | Transmission (0 = automatic, 1 = manual)
gear  | Number of forward gear
carb  | Number of carburetors

The variables are all numeric.

**Solution**  
After conducting some exploratory analysis, we have fit different linear regression models and we have discovered that manual transmission cars are better for mpg than automatic ones but there are other variables like weight and cylinder that are better predictor for MPG and the transmission type depends on them.

### Solution details

**Load the data**  
After loading the data, the transmission type (am) is transformed in a factor for convenience.
```{r}
data(mtcars)
mtcars$am = factor(mtcars$am, label = c("automatic", "manual"))
```
**Exploratory analysis**  
From the summary below and the boxplot in the appendix (figure 1), it is possible to see that MPG statistics (mean, median, ...) are greater for manual transmission than automatic transmission. This suggests that manual transmission could be better than automatic transmission for MPG.
```{r}
tapply(mtcars$mpg, mtcars$am, summary)
```
**Fit multiple models and model selection**  
The first model considered is the simplest model with transmission type (am) as the only one explanatory variable. 
```{r}
fit.simple <- lm(mpg ~ am, data = mtcars)
summary(fit.simple)$coeff
```
Interpreting the coefficients of the summary above, the model estimates an expected increase of `r round(summary(fit.simple)$coeff[2,1],3)` MPG for a change from automatic transmission to manual transmission, on average.

After calculating confidence intervals, 
```{r}
confint(fit.simple)
```
it is possible to affirm with 95% confidence that a change from automatic transmission to manual transmission results in a `r round(confint(fit.simple)[2,1],3)` to `r round(confint(fit.simple)[2,2],2)` increase in MPG.
Nevertheless, this model is not very expressive because the adjusted R-squared is `r summary(fit.simple)$adj.r.squared` and this means that the model explains only `r round(summary(fit.simple)$adj.r.squared*100,2)`% of the variance. We use adjusted R-squared instead of R-squared to measure the regression model because adjusted R-squared does not increase if we add variables that doesn't really provide any new information.
A general approach to find a better model is to consider the most complete model with all the variables of the data set as explanatory variables and then applying stepwise model selection with backwards elimination technique.     

Before fitting the model, we transform the numeric variables cyl, vs, gear and carb in factors for convenience because they can assume only few integer values.
```{r}
mtcars$vs = factor(mtcars$vs)
mtcars$cyl = factor(mtcars$cyl)
mtcars$gear = factor(mtcars$gear)
mtcars$carb = factor(mtcars$carb)
```
**Final model**  
The explanatory variables of the final model are cylinders (cyl), gross horsepower (hp), weight (wt) and transmission type (am). 

```{r, results='hide'}
fit.total <- lm(mpg ~ ., data = mtcars)
fit.final <- step(fit.total)
```
```{r}
summary(fit.final)$coeff
```
The adjusted R-squared is now `r summary(fit.final)$adj.r.squared` so this model is more expressive than the first one. 

Holding the remaining variables constant, the new model estimates an expected increase of `r round(summary(fit.final)$coeff[6,1],3)` MPG for a change from automatic transmission to manual transmission, on average . Therefore transmission type appears to have a lower impact on MPG than if all the other variables are disregarded as in the first model.
Moreover, the transmission type is not statistically significant for this model if we use 0.05 as a type I error rate significance benchmark (the p-value of transmission type is 0.2065 and so larger than 0.05). In the new model indeed, the most significant variable is the weight and the model estimates an expected decrease of `r abs(round(summary(fit.final)$coeff[5,1],3))` MPG for every one ton increase in weight, holding the remaining variables constant.

After calculating confidence intervals, 
```{r}
confint(fit.final)
```
it is possible to affirm with 95% confidence that an increase of one ton in weight results in a `r abs(confint(fit.final)[5,2])` to `r abs(confint(fit.final)[5,1])` decrease in MPG.
To summarize, manual transmission cars have higher MPG but probably due to their weight (and number of cylinders, not discussed here for reason of space).
 
**Residual diagnostic**  
The residuals diagnostic plots in the appendix (figure 2) confirm a good regression fit. In more details, in the Residuals vs Fitted plot the points are randomly scattered with no particular pattern and in the Normal Q-Q plot the points follow the line so it is possible to conclude that the residuals are normally distributed around 0 with constant variability. In the other two plots, the points are more or less grouped near the center.   
To summarize, the conditions for applying the final linear regression model are satisfied. 
 

------------


### Appendix

**Figure 1. Relationship between MPG and transmission type**
```{r, fig.width = 5, fig.height = 5}
boxplot(mtcars$mpg ~ mtcars$am, xlab = "Trasmission type", ylab = "MPG Miles/(US) gallon")
```   

**Figure 2. Residual diagnostics**
```{r, fig.height = 10, fig.width = 10}
par(mfrow = c(2, 2))
plot(fit.final)
```


