# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

1. Load the data

    ```{r}
    activity <- read.csv("activity.csv")
    ```



## What is mean total number of steps taken per day?

1. Histogram of the total number of steps taken each day

    ```{r fig.width = 15}
    steps.date <- aggregate(steps ~ date, data = activity, FUN = sum)
    barplot(steps.date$steps, names.arg = steps.date$date, col = "blueviolet", xlab  
    = "", ylab = "steps", main = "Total number of steps per day", las = 2)
    ```
    
2. **Mean** and **median** total number of steps taken per day

    ```{r}
    mean(steps.date$steps)
    median(steps.date$steps)
    ```

## What is the average daily activity pattern?

1. Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days
 
    ```{r}
    steps.interval <- aggregate(steps ~ interval, data = activity, FUN = mean)
    plot(steps.interval, type = "l", main = "Average number of steps across all days")
    ```

2. The 5-minute interval that, on average across all the days in the dataset, contains the maximum number of steps

    ```{r}
    steps.interval$interval[which.max(steps.interval$steps)]
    ```


## Imputing missing values

There are a number of days/intervals where there are missing
values (coded as `NA`). The presence of missing days may introduce
bias into some calculations or summaries of the data.

1. Total number of missing values in the dataset

    ```{r}
    sum(is.na(activity$steps))
    ```

2. Fill the missing values in the dataset with the mean for that 5-minute interval and create a new dataset equal to the original one but with the missing data filled in

    ```{r}
    colnames(steps.interval)[2] <- "steps.mean"   
    activity <- merge(activity, steps.interval, by = "interval")

    na.index <- is.na(activity$steps)
    activity$steps[na.index] <- activity$steps.mean[na.index]

    activity <- activity[, 1:3]
    ```

3. Histogram of the total number of steps taken each day with missing values filled in 

    ```{r fig.width = 15}
    steps.date <- aggregate(steps ~ date, data = activity, FUN = sum)
    barplot(steps.date$steps, names.arg = steps.date$date, col = "blueviolet", xlab = "", ylab     
    = "steps", main = "Total number of steps per day", las = 2)
    ```

4. **Mean** and **median** total number of steps taken per day with missing values filled in

    ```{r}
    mean(steps.date$steps)
    median(steps.date$steps)
    ```

## Are there differences in activity patterns between weekdays and weekends?

1. New factor variable with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

    ```{r results = "hide",  message = FALSE, warning = FALSE}
    Sys.setlocale(locale = "C")
    install.packages("ggplot2", repos="http://cran.rstudio.com/")
    library(ggplot2)
    ```
    
    ```{r}
    weekend <- weekdays(as.Date(activity$date)) %in% c("Saturday", "Sunday")
    activity$weekdays[weekend] <- "weekend"   
    activity$weekdays[!weekend] <- "weekday"   
    activity$weekdays <- factor(activity$weekdays)
    ```

2. Panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days 

    ```{r}
    weekend = subset(activity, weekdays == "weekend")
    steps.interval.weekend <- aggregate(steps ~ interval, data = weekend, FUN = mean)
    steps.interval.weekend$weekdays = "weekend"

    weekday = subset(activity, weekdays == "weekday")
    steps.interval.weekday <- aggregate(steps ~ interval, data = weekday, FUN = mean)
    steps.interval.weekday$weekdays = "weekday"

    steps.interval <- rbind(steps.interval.weekend, steps.interval.weekday)

    qplot(interval, steps, data = steps.interval, facets = weekdays ~ ., geom = "line", 
          ylab = "Number of   steps")
    ```

